---
title: "PM21134 - ChIP of SWI/SNF and ASCL1"
author: "Stephanie Strohbuecker"
subtitle: Read alignment using the nf-core ChIP seq pipeline
output:
  html_document:
    df_print: paged
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      eval.after = "fig.cap",
                      fig.width = 12)
```

We are using the [nf-core ChIP pipeline (version 1.2.2)](https://nf-co.re/chipseq/1.2.2) for aligning the reads from the ChIP experiment.


```{r}
RMD_file_name <- "0-2.read_alignment_nfcore"
```

## Libraries

```{r}
library(tidyverse)
library(here)
```

```{r}
pkgload::load_all() # load all the functions available in this project
```
The data for the experiment is associated with two sequencing runs:

* 220218_A01366_0146_BH757VDMXY
* 220325_A01366_0163_AH3CWHDSX3

Note that the second sequencing run data was generated in paired-end mode. However, as the first sequencing run was single-end, we only used the `R1` information as input.

Generate the experimental table input necessary for input into the nf-core ChIP-seq pipeline:
```{r}
sample_info <-
  read_csv(here("data", "raw_data", "project_docs", "PM21134_design.csv")) %>%
  rename_with(~ str_replace_all(., " ", "_")) %>%
  select(sample_lims, sample_name, replicate, fastq_1) %>%
  separate(
    sample_name,
    into = c("antibody", "genotype"),
    sep = "_ChIP_",
    remove = FALSE
  )  %>%
  mutate(
    genotype = str_remove(genotype, "Input_") %>%
      str_remove("\\d$") %>%
      str_remove("-.*$"),
    group = case_when(
      str_detect(sample_name, "Input") ~ str_c(genotype, antibody, "Input", sep = "_"),!is.na(antibody) ~ str_c(genotype, antibody, sep = "_")
    ),
    replicate = str_remove(replicate, "^R"),
    antibody = case_when(
      str_detect(sample_name, "Input") ~ NA_character_,
      TRUE ~ antibody
    ),
    control = case_when(
      !is.na(antibody) ~ str_c(group, "Input", sep = "_"),
      TRUE ~ NA_character_
    ),
    fastq_2 = NA_character_
  ) %>%
  select(group,
         replicate,
         fastq_1,
         fastq_2,
         antibody,
         control,
         everything()) %>%
  arrange(antibody, group, replicate)

```

Note that the ChIP for SMARCA1 failed. Therefore Oana Paun would like to use an older ChIP for SMARCB1 that was run previously as part of the following project/ run:

* DN19154/ 190924_K00102_0399_BHCFW3BBXY

```{r}
DN19154_sample_info <- readxl::read_excel(
  here(
    "data",
    "raw_data",
    "project_docs",
    "SMARCB1_ChIP_replacment_info_vs09-05-2022.xlsx"
  ),
  skip = 1
) %>%
  select(sample_lims = `Sample limsid`, sample_name = `Sample Name`) %>%
  arrange(sample_name) %>%
  mutate(
    group = str_remove(sample_name, "1percent_") %>%
      str_remove("ChIP_") %>%
      str_remove("_$") %>%
      str_remove("_K4.*$|_C1.*$") %>%
      str_remove("ASCL1_")
  ) %>%
  group_by(group) %>%
  mutate(replicate = row_number() %>% as.character()) %>%
  ungroup() %>%
  separate(group, into = c("antibody", "genotype")) %>%
  mutate(
    genotype = str_replace(genotype, "KO", "ASCL1_KO"),
    group = str_c(antibody, genotype, sep = "_"),
    fastq_2 = NA_character_,
    antibody = case_when(
      str_detect(sample_name, "INPUT") ~ NA_character_,
      TRUE ~ antibody
    ),
    control = case_when(
      !is.na(antibody) ~ str_c("INPUT", genotype, sep = "_"),
      TRUE ~ NA_character_
    )
  )

sample_info <- read_csv(here("data",
              "raw_data",
              "project_docs", "DN19154_design.csv")) %>%
  select(fastq_1) %>%
  mutate(sample_lims = basename(fastq_1) %>%
           str_remove("_.*$")) %>%
  inner_join(DN19154_sample_info, by = "sample_lims") %>%
  # select(sample_name) %>% distinct() %>%
  arrange(group, replicate) %>%
  select(
    group,
    replicate,
    fastq_1,
    fastq_2,
    antibody,
    control,
    sample_lims,
    sample_name,
    genotype
  ) %>% 
  bind_rows(sample_info)
```


```{r}
sample_info %>%
  write_csv(str_c(
    here("data", "raw_data", "project_docs"),
    "PM21134_sample_information.csv",
    sep = "/"
  ))
```

To run read alignment using the ChIP nf-core pipeline, we need to prepare a samplesheet with the following columns:

* group
* replicate
* fastq read 1
* fastq read 2 (if applicable)
* antibody
* control

```{r}
generate_directories(str_c(here(), "data", "derived_data", "nfcore_ChIP", sep = "/"))
```


```{r}
sample_info %>%
  select(group, replicate, fastq_1, fastq_2, antibody, control) %>%
  write_csv(
    file = str_c(
      here(),
      "data",
      "derived_data",
      "nfcore_ChIP",
      "design.csv",
      sep = "/"
    ),
    na = ""
  )
```

The results for the nf-core RNA-seq pipeline will be stored here:
```{r}
str_c(here(), "data", "derived_data", "nfcore_ChIP", sep = "/")
```

  
```{r}
log_dir <- here("analysis", "logs")
scripts_dir <- here("analysis", "scripts")
```


The `custom.config` is empty (increased memory for bwa due to error during the pipeline run)
```{r}
write_lines(
  "process {
  withName:bwa_align {
    memory = 64.GB
  }
}",
file = str_c(
  here(),
  "data",
  "derived_data",
  "nfcore_ChIP",
  "custom.config",
  sep = "/"
)
)
```

(Note that the slurm command needs to be run from the following directory: `/camp/stp/babs/working/strohbs/projects/guillemotf/oana.paun/op17_PM21134_ChIP_SWI_SNFxASCL1/data/derived_data/nfcore_ChIP`).

Note we ran the pipeline twice once with `--broad` and once with `--narrow_peak` option (note that using `-resume` avoids running the whole pipeline).
```{r}
slurm_cmd <-
  str_c(
    "sbatch -c 1 --mem-per-cpu 7G -N 1 --time 16:00:00 --output",
    str_c(log_dir, "nf_core_ChIP.log", sep = "/"),
    "--error",
    str_c(log_dir, "nf_core_ChIP_error.log", sep = "/"),
    str_c(scripts_dir, "0-2a.Run_nfcore_ChIP.sh", sep = "/"),
    sep = " "
  ) 

slurm_cmd
```
